---
title: "End-to-end MLE workflow with xsdm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{End-to-end MLE workflow with xsdm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
set.seed(123)
```

## Overview

`xsdm` reconstructs a species' fundamental ecological niche by maximising
a likelihood that combines:

* an **environmental time-series** (locations × time × variables),
* **presence / absence** records on the same locations.

This vignette walks through the complete workflow on the bundled
`example_1` fixture: build a starting design, run multi-start MLE, inspect
results, and project them spatially.

## Setup

```{r setup}
library(xsdm)

# The example_1 object ships with the package and provides a small,
# fully-formed environmental array + occurrence vector.
str(example_1, max.level = 1)
```

## 1. Multi-start maximum-likelihood estimation

`optimize_likelihood()` is the user-facing entry point. It builds a
Sobol' multi-start design of `num_starts` parameter vectors over a
data-driven range, runs each start through the C++ engine, and returns
both the per-start solutions and the best one.

```{r optimize}
fit <- optimize_likelihood(
  env_dat     = example_1$env_array,
  occ         = example_1$occ_vec,
  num_starts  = 8L,        # use 50-200 for real analyses
  parallel    = FALSE,     # set TRUE on multi-core machines
  control     = list(maxeval = 200),
  verbose     = FALSE
)

# Best solution and its log-likelihood
fit$best$loglik
fit$best$full_par
```

## 2. Habitat-suitability projection

```{r vsp, eval = FALSE}
# The `vsp` function projects the best-fit niche onto a raster grid.
# Skipped at vignette-build time because terra requires a raster on disk.
hs <- vsp(
  param_list = interpret_parameters(fit$best$full_par,
                                    p = dim(example_1$env_array)[3]),
  env_layers = list(example_1$bio01, example_1$bio12)
)
```

## 3. Profile likelihood

```{r profile, eval = FALSE}
# Profile the detection probability `pd` around the MLE.
prof <- profile_likelihood(
  env_dat = example_1$env_array,
  occ     = example_1$occ_vec,
  param   = "pd",
  best    = fit$best$full_par,
  n_grid  = 7,
  span    = 0.5
)
```

## See also

* `vignette("xsdm")` — package overview *(TODO)*.
* `?optimize_likelihood` — full argument reference.
* `?profile_likelihood` — uncertainty estimation.
* `?make_mask_names` — fixing specific parameters during optimization.
