---
title: "Unusable models: when a model does not have a maximum likelihood"
author:
  - name: "Daniel C. Reuman"
    affiliation: "Department of Ecology and Evolutionary Biology and Center for Ecological Research, University of Kansas"
  - name: "Angel Luís Robles Fernández"
    affiliation: "Department of Ecology and Evolutionary Biology and Center for Ecological Research, University of Kansas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Unusable models: when a model does not have a maximum likelihood}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<div style="display:none">
$$
\newcommand{\mean}[1]{\overline{#1}}
\newcommand{\var}{\text{var}}
\newcommand{\cov}{\text{cov}}
\newcommand{\cor}{\text{cor}}
\newcommand{\Rp}{\text{Re}}
\newcommand{\E}{\text{E}}
\newcommand{\ltsgr}{\text{ltsgr}}
\newcommand{\expit}{\text{expit}}
\newcommand{\logit}{\text{logit}}
$$
</div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=TRUE,
  message=FALSE,
  warning=FALSE,
  cache=FALSE,
  eval=FALSE
)
```

**Abstract.** When an xsdm model does not have a maximum likelihood, it should not be used.
This document shows an example of how that can occur, and how to diagnose it.
Along the way, another type of boundary model (in addition to the $p_d=1$ case
described elsewhere) is
illustrated, where sigma parameters are set to infinity, corresponding to
insensitivity of annual net growth to environmental changes in a certain direction
in environment space.

The Eastern glass lizard, *Ophisaurus ventralis*, is a legless lizard
found in the southeastern United States. It is the longest and heaviest species
of its genus, growing up to 108cm in total length.

We start by loading the data:
```{r bring_in_the_data}
library(xsdm)
env_array <-  example_2$env_array

#We divide by 100 to get the correct units in the example
env_array <-  example_2$env_array/100
dim(env_array)
dimnames(env_array)[[3]]
occ <-  example_2$occ_vec
length(occ)
```

Here, there are 6 environmental variables recorded for
39 years in 2728 locations, with
accompanying detections and pseudo-absences in the variable `occ`.
The first three environmental variables (BIO1, BIO10, BIO11) are temperature
variables, and the last three (BIO12, BIO16, BIO17) are precipitation variables.
BIO1 is mean annual temperature, BIO10 is mean temperature of the warmest quarter,
BIO11 is mean temperature of the coldest quarter, BIO12 is annual precipitation, BIO16 is
precipitation of the wettest quarter, and BIO17 is precipitation of the driest quarter.

Now look at the distributions of values of environmental variables to make sure they
are not on very different scales, which would cause problems for optimization:
```{r look at distributions}
apply(FUN=quantile, X=env_array, MARGIN=3,prob=c(.025,.25,.5,.75,.975))
```

These distributions look basically OK.

Now fit 15 models, each from 25 starting conditions:
```{r fit_models,cache=TRUE}
models <- matrix(c(1,0,0,0,0,0,
                  0,1,0,0,0,0,
                  0,0,1,0,0,0,
                  0,0,0,1,0,0,
                  0,0,0,0,1,0,
                  0,0,0,0,0,1,
                  1,0,0,1,0,0,
                  1,0,0,0,1,0,
                  1,0,0,0,0,1,
                  0,1,0,1,0,0,
                  0,1,0,0,1,0,
                  0,1,0,0,0,1,
                  0,0,1,1,0,0,
                  0,0,1,0,1,0,
                  0,0,1,0,0,1), nrow=15, byrow=TRUE)
all_model_results <- list()
for (i in 1:nrow(models))
{
  print(i)
  env_dat <-  env_array[ , , models[i,]==1, drop = FALSE]
  starts <-  xsdm::start_parms(env_dat,num_starts=25)
  all_optim_results <-  list()
  for (j in 1:nrow(starts))
  {
    all_optim_results[[j]] <-  optim(par = starts[j,],
                                   fn = xsdm::loglik_math,
                                   method = "BFGS",
                                   env_dat = env_dat,
                                   occ = occ,
                                   negative=TRUE,
                                   control = list(trace=0)
                                   )
  }
  all_model_results[[i]] <- all_optim_results
}
```

Within each model, rank the optimization results:
```{r rank_optim_results}
for (i in 1:length(all_model_results))
{
  values <-  sapply(X = all_model_results[[i]], FUN=function(x){x$value})
  inds <-  order(values)
  all_model_results[[i]] <-  all_model_results[[i]][inds]
}
```

Rank the models by BIC, bearing in mind that we've been working with the
negative of the likelihood:
```{r rank_by_BIC}
model_BICs <- sapply(X=all_model_results,
                      FUN=function(x){
                        best_loglik = x[[1]]$value
                        num_parms = length(x[[1]]$par)
                        n = length(occ)
                        BIC = 2*best_loglik + num_parms*log(n)
                        return(BIC)
                      }
                    )
```

Also by AIC, then display:
```{r rank_by_AIC}
model_AICs <- sapply(X=all_model_results,
                      FUN=function(x){
                        best_loglik = x[[1]]$value
                        num_parms = length(x[[1]]$par)
                        AIC = 2*best_loglik + 2*num_parms
                        return(AIC)
                      }
                    )

inds <- order(model_BICs)
rbind(model_BICs[inds],model_AICs[inds])
plot(model_BICs,model_AICs,type="p",xlab="BIC",ylab="AIC")
order(model_BICs)
order(model_AICs)
```

The AIC and BIC results are pretty well aligned, and the four
best models are the same:
```{r look_at_models}
models[order(model_BICs)[1:4],]
```

These four models all use the same predictor, the fifth one, and
then some use various possible second predictors.

We emphasize that these BIC and AIC values may or may not be meaningful, since
B/AIC can only be computed when the likelihood has been effectively maximized.
We will elaborate below, but for now we accept these are pseudo-BIC and
pseudo-AIC values, subject to later validation or rejection.

Among the models we fitted, there is one clear winner in pseudo-BIC (the model
with the lowest pseudo-BIC). So let's investigate it further.
Start by optimizing it a bit harder to see if we can
do any better.
```{r check_model_11,cache=TRUE}
i <- 11
env_dat <- env_array[,,models[i,]==1,drop=FALSE]
starts <- xsdm::start_parms(env_dat, num_starts = 100)
model_11_results <- list()
for (j in 1:nrow(starts))
{
  model_11_results[[j]] <- optim(par=starts[j,],fn=xsdm::loglik_math,
                                method="BFGS",
                                env_dat = env_dat,
                                occ = occ,negative=TRUE,
                                control = list(trace=0))
}
all_model_results[[11]][[1]]$value
min(sapply(X=model_11_results, FUN=function(y){y$value}))
```

We did slightly better.

Now move forward by looking at the results for this model, starting by
writing a convenience function for examine optimization results:
```{r look_at_optim_results}
examine_optim_results <- function(optim_results,mask=NULL)
{
  #put optimization results in order from best to worst
  bestlogliks <- sapply(X=optim_results,FUN=function(x){x$value})
  inds <- order(bestlogliks)
  bestlogliks <- bestlogliks[inds]
  optim_results <- optim_results[inds]

  #model convergence
  convergences <- sapply(X=optim_results,FUN=function(x){x$convergence})

  #compute distances to the best result in parameter space
  best_parms_math <- optim_results[[1]]$par
  parms_dists_to_best <- lapply(
    X=optim_results,
    FUN=function(x){
      xsdm::dist_between_params(
        x$par,
        best_parms_math,
        mask=mask,
        give_closest_rep=TRUE)
    }
  )
  parms_dists <- sapply(X=parms_dists_to_best, FUN=function(x){x$distance})
  bestparms <- sapply(X=parms_dists_to_best, FUN=function(x){unlist(x$representative)})

  #put it all together
  return(rbind(bestlogliks,convergences,parms_dists,bestparms))
}

h <- examine_optim_results(all_model_results[[11]])
t(h[ ,1:8])
```

The very large values of `sigltil2` suggest the boundary
model where this parameter is set to `Inf`, corresponding to a direction
in environment space along which annual net growth is insensitive to changes in
the environment.

So we consider the corresponding boundary model:
```{r fit_boundary_model,cache=TRUE}
i <- 11
env_dat <- env_array[ , , models[i,] == 1, drop=FALSE]
mask <- c(sigltil2 = Inf)
new_starts <- xsdm::start_parms(env_dat[occ == 1, , , drop=FALSE],
                               mask = mask,
                               num_starts = 100)

bdry_optim_results <- list()
for (j in 1:nrow(new_starts))
{
  bdry_optim_results[[j]] <- optim(par = new_starts[j,],
                                  fn = xsdm::loglik_math,
                                  method = "BFGS",
                                  env_dat = env_dat,
                                  occ = occ,
                                  mask = mask,
                                  negative = TRUE,
                                  control = list(trace=0, maxit=500))
}
```

Now look at these results:
```{r look_at_bdry_models}
h <- examine_optim_results(bdry_optim_results, mask = mask)
t(h[ ,1:8])
```

The best likelihoods obtained for this boundary model are very similar
to those obtained for the initial, non-boundary model, and the boundary model has one
fewer parameter, so we tentatively adopt the boundary model over the
earlier model. However, these optimization results reveal that we probably still
have not successfully optimized the likelihood, or that the  likelihood
surface may have a ridge or an asymptote or other pathological feature.
One sees this by observing that whereas the top several optimization results
are similar in likelihood, they are spread out in parameter space (the
`parms\_dists` column shows distance in parameter space to the
top-likelihood result).

To investigate the model further, we profile:
```{r profile, cache=TRUE}
pnames <-  names(xsdm::make_mask_names(2))
pnames <-  pnames[!(pnames %in% names(mask))]

values <-  sapply(X=bdry_optim_results, FUN=function(x){x$value})
inds <-  order(values)
bdry_optim_results <-  bdry_optim_results[inds]

all_profiles <- list()
linc <-  rep(0.05, length(pnames))
rinc <-  rep(0.05, length(pnames))
linc[8] <-  0.001
rinc[8] <-  0.001
for (counter in 1:length(pnames))
{
  all_profiles[[counter]] <-  xsdm::profile_likelihood(
                              profile_parameter = pnames[counter],
                              increment_left =linc[counter],
                              increment_right = rinc[counter],
                              num_steps_left = 50,
                              num_steps_right = 50,
                              alpha = 0.95,
                              optim_param_vector = bdry_optim_results[[1]]$par,
                              env_dat=env_dat,
                              occ = occ,
                              mask = mask,
                              num_threads = 6
                            )
}
names(all_profiles) <-  pnames
```

Now plot these profiles:
```{r plot_profiles}
plot_tool <- function(ap,index)
  {
  x <- ap[[index]]$profile$value_math
  y <- ap[[index]]$profile$loglik
  xlab <- names(ap)[index]
  thresh <- ap[[index]]$threshold
  plot(x,y,
       type="o",xlab=xlab,
       ylab="Log likelihood")
  lines(range(x),rep(thresh,2),type="l",
        lty="dashed",col="red")
}

par(mfrow=c(3,3))
plot_tool(all_profiles, 1)
plot_tool(all_profiles, 2)
plot_tool(all_profiles, 3)
plot_tool(all_profiles, 4)
plot_tool(all_profiles, 5)
plot_tool(all_profiles, 6)
plot_tool(all_profiles, 7)
plot_tool(all_profiles, 8)
```

These profiles are not dome-shaped, and have other idiosyncratic features,
confirming that we had not effectively optimized the likelihood.
The `mu2`, `mu2`, and `ctil` profile, in particular, show problems.

We do a pairs plot based on the `mu2` profile to investigate further:
```{r make_pairs_plot_again}
p <- all_profiles[[2]]$parameters
head(p)
p <- p[,1:8]
pairs(p)
```

This pairs plot helps us see the tradeoff going on
between parameters.
As `mu2` is increased, other parameters, especially `ctil`,
change monotonically. These results suggest a ridge in the
likelihood surface that appears to rise asymptotically along some path
in parameter space for which `mu2` is increasing. We have not identified
a maximum of the likelihood function, and it looks as though there may not
be one if the increase is indeed asymptotic.

We look at what the growth-environment function looks like
for the best parameters we have found so far, maybe that will give some insight into what
is going wrong:
```{r look_at_inferred_ge_func,fig.height=4}
param_list <- bdry_optim_results[[1]]$par
param_list["sigltil2"] <- Inf
param_list <- param_list[names(xsdm::make_mask_names(2))]
param_list_bio <- xsdm::math_to_bio(param_list)
xsdm::interpret_parameters(param_list = param_list_bio,
                          plot_indices = c(1,2), env_dat=env_dat, occ = occ)
param_list_bio$mu
```

Estimated values of `mu2` are outside
the range of the environmental data. The actual distribution of the species is across
Florida and along the coasts of the Southeastern United States,
and the southern extent of the
species is very likely constrained, on the southern end of the range,
by the Atlantic Ocean and the Gulf of Mexico
rather than by temperature and precipitation.
Ultimately the modeling probably fails here for these reasons.
A Bayesian approach to the same problem could likely address some of the
limitations, here, by setting appropriate priors. In the frequentist setting,
one must discard the model. AIC and BIC values are invalid when the
likelihood has not been adequately optimized and when the likelihood
surface in the vicinity of the optimum is not approximately a dome.

Immediate next steps should include examination of the
second- and third-best models according to pseudo-BIC, to see if they have similar problems.
For brevity and because we are just trying to illustrate statistical principles and
workflows, we skip straight to examining the fourth-best model,
which may be simpler because it only uses one environmental variable.
```{r examine_result_model_5}
i <- 5
model_5_optim_results <-  all_model_results[[i]]
h <- examine_optim_results(model_5_optim_results)
t(h[,1:8])
```

One can see the boundary model with `sigrtil` set to `Inf`
should be considered:
```{r fit_boundary_model_5,cache=TRUE}
env_dat <- env_array[,,models[i,]==1,drop=FALSE]
mask <- c(sigrtil1 = Inf)
new_starts <- xsdm::start_parms(env_dat[occ==1,,,drop=FALSE],mask=mask,
                               num_starts=100)

bdry_optim_results5 <- list()
for (j in 1:nrow(new_starts))
{
  bdry_optim_results5[[j]] <- optim(par=new_starts[j,],fn=xsdm::loglik_math,
                                method="BFGS",
                                env_dat=env_dat,occ=occ,mask=mask,negative=TRUE,
                                control=list(trace=0,maxit=500))
}
```

Examine these results:
```{r look_at_bdry_models_5}
h <- examine_optim_results(bdry_optim_results5,mask=mask)
t(h[,1:8])
values <- sapply(X=bdry_optim_results5, FUN=function(x){x$value})
inds <- order(values)
bdry_optim_results5 <- bdry_optim_results5[inds]
```

This model looks like it probably optimized well.

Next do profiles:
```{r profile_model_5, cache=TRUE}
all_profiles <- list()
pnames <- names(xsdm::make_mask_names(1))
pnames <- pnames[pnames!="sigrtil1"]
linc <- c(0.05,0.025,0.05,0.05)
rinc <- c(0.15,0.025,0.05,0.05)
for (counter in 1:length(pnames))
{
  all_profiles[[counter]] <-  xsdm::profile_likelihood(
                              profile_parameter = pnames[counter],
                              increment_left  = linc[counter],
                              increment_right = rinc[counter],
                              num_steps_left  = 50,
                              num_steps_right = 50,
                              alpha = 0.95,
                              optim_param_vector = bdry_optim_results5[[1]]$par,
                              env_dat = env_dat,
                              occ = occ,
                              mask = mask,
                              num_threads = 6
                            )
}
names(all_profiles) <- pnames

par(mfrow=c(2,3))
plot_tool(all_profiles,1)
plot_tool(all_profiles,2)
plot_tool(all_profiles,3)
plot_tool(all_profiles,4)
```

The profiles look adequate. So the model is suitable for use, unlike
the previous model. The BIC and AIC for this model are valid. If the second- or
third-best model is suitable in the same manner, the lowest-BIC model
suitable model should tend to be preferred.
